sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-01-25 doc/paper-jss/sdmTMB-cover.tex
Update sdmTMB-cover.tex
Feel free to take or leave these -- mostly trying to shorten / focus

We believe that \pkg{sdmTMB} represents a major advance in terms of speed, functionality, and accessibility for fitting SPDE (stochastic partial differential equation)-based GLMMs to spatial or spatiotemporal data.

@@

Functionality: This package represents a spatial analog of the \pkg{glmmTMB} \proglang{R} package, with additional features beyond random fields (penalized smoothers as in \pkg{mgcv}, priors, threshold/breakpoint models) and spatial anisotropy (directionally dependent spatial correlation; lacking in \pkg{INLA} and \pkg{inlabru}). Spatial anisotropy is critical in many applied contexts such as when modelling species distributions near any geographic boundary or along an environmental gradient.

Accessibility: \pkg{sdmTMB} is a much more usable interface than alternatives for applied users, who may be familiar with \proglang{R}'s \texttt{glm()}, \pkg{glmmTMB}, \pkg{lme4}, or \pkg{mgcv}. In our experience, such applied users struggle to code these types of models in \pkg{INLA}; \pkg{inlabru} improves the experience but still requires an understanding of \pkg{INLA} to fit many useful but complex structures. Furthermore, \pkg{INLA} and \pkg{inlabru} are fundamentally Bayesian and require users to deal with the complexity of priors. \pkg{sdmTMB} can incorporate priors and extend to full Bayesian inference via \pkg{Stan}, while by default allowing users the simplicity and speed of maximum likelihood.

As evidence of its utility, there are already \href{https://github.com/pbs-assess/sdmTMB/wiki/Publications-using-sdmTMB}{16 published preprints, papers, or reports} using \pkg{sdmTMB}, six funded research grants, an ever-growing number of ongoing research projects, and workshops that in total have been attended by $>$200 participants.

>>>>

We believe that \pkg{sdmTMB} represents a major advance in terms of functionality, accessibility, and speed for fitting the broad class of SPDE (stochastic partial differential equation)-based GLMMs to spatial or spatiotemporal data.

Functionality: This package represents a spatial analog of the \pkg{glmmTMB} \proglang{R} package, with additional features beyond random fields (penalized smoothers as in \pkg{mgcv}, priors as in \pkg{INLA}, threshold/breakpoint models), spatial anisotropy (directionally dependent spatial correlation), and methods for model comparison (cross-validation, stacking).

Accessibility: \pkg{sdmTMB} provides a much more usable interface than alternatives, particular for users who may be familiar with the syntax found in \proglang{R}'s \texttt{glm()}, \pkg{glmmTMB}, \pkg{lme4}, or \pkg{mgcv} packages. Existing packages, such as \pkg{INLA} or \pkg{inlabru} are fundamentally Bayesian and require users to deal with the complexity of priors. While \pkg{sdmTMB} can incorporate priors and extend to full Bayesian inference via \pkg{Stan}, by default estimation is done using the simplicity and speed of maximum likelihood.

@@

As evidence of its utility, there are already \href{https://github.com/pbs-assess/sdmTMB/wiki/Publications-using-sdmTMB}{16 published preprints, papers, or reports} using \pkg{sdmTMB}, six funded research grants, an ever-growing number of ongoing research projects, and international workshops that in total have been attended by $>$200 participants.
